\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-10-26 paper/main.tex
edits up through pg 4 (fig 1 caption)

  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments, to containers, to virtual
  machines.  Each of these existing solutions requires installing or
  setting up a system for running the desired code, increasing the 
  complexity and time cost of sharing or engaging with reproducible 
  science. Here, we propose a lighter-weight solution:  the
  \texttt{davos} library.  When used in combination with a
  notebook-based Python project, the \texttt{davos} library provides a
  mechanism for specifying (and automatically installing) the correct
  versions of the project's dependencies.  The \texttt{davos}
  library further ensures that those packages and specific versions are used every time the
  notebook's code is executed.  This enables researchers to share a complete
  reproducible copy of their code within a single Jupyter notebook file.

@@

Within the Python~\cite{vanR95} community, external
packages that are published in the most popular
repositories~\cite{Pyth03, cond15} are associated with version numbers
and tags that allow users to guarantee they are installing
exactly the same code across different computing environments \cite{CoghStuf13}.
Despite that it is \textit{possible} to manually install the intended
version of every dependency of a Python script or package, manually
tracking down those dependencies can impose a substantial burden on 
the user and make room for mistakes. Further, when dependency versions 
are left unspecified, replicating the original computing environment 
becomes difficult or impossible.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code.  Scripts may reference external libraries, but
  those libraries must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those libraries were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.).  When the \textbf{\texttt{davos} library} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install the required external libraries (at their
  correct versions, when specified).  \textbf{Virtual environments}
  install an isolated copy of Python and all required dependencies.
  This typically requires defining a \texttt{requirements.txt} file or
  an environment (\texttt{.yml}) file that that specifies all project
  dependencies (including version numbers of external libraries).
  \textbf{Containers} provide a means of defining an isolated
  environment that includes a complete operating system (independent
  of the user's operating system), in addition to (optionally)
  specifying a virtual environment or other configurations needed to
  provide the necessary computing environment.  Containers are
  typically defined using specification files (e.g., a plain-text
  \texttt{Dockerfile}) that instruct the virtualization engine
  regarding how to build the virtual environment.  \textbf{Virtual
    machines} provide a complete hardware-level simulation of the
  computing environment.  In addition to simulating specific hardware,
  virtual machines (typically specified using binary images files)
  must also define operating system-level properties of the computing
  environment.  Systems to the left of the blue vertical line entail
  sharing individual files, with no additional installation or
  configuration needed to run the target code.  Systems to the right
  of the red vertical line support precise control over dependencies
  and versioning.  Notebooks enhanced using the \texttt{davos} library
  are easily shareable and require minimal setup costs, while also
  facilitating high reproducibility by enabling precise control over
  project dependencies.}
\label{fig:code-sharing}
\end{figure}

Computational researchers and other programmers have developed a broad
set of approaches and tools to facilitate code sharing and
reproducible outcomes (Fig.~\ref{fig:code-sharing}). At one
extreme, simply distributing a set of Python scripts (\texttt{.py} files) may
enable others to use or gain insights into the relevant work. Because
Python is installed by default on most modern operating systems, for
some projects, this may be sufficient. Another popular approach
entails creating Jupyter
notebooks~\cite{KluyEtal16} that comprise a mix of text, executable
code, and embedded media. Notebooks may call or import external
scripts or libraries---even intersperse snippets of other programming 
or markup languages---in order to provide a more compact and readable
experience for users. Each of these systems (Python scripts and
notebooks) provides a convenient means of sharing code, with the
caveat that they do not specify the computing environment in which the
code is executed. Therefore the functionality of code shared using
these systems cannot be guaranteed across different users or setups.

At another extreme, virtual machines~\cite{Gold74, AltiEtal05,
 Rose99} provide a hardware-level simulation of the desired system.
Virtual machines are typically isolated such
that installing or running software on a virtual machine does not
impact the user's primary operating system or computing environment.
Containers~\cite[e.g.,][]{Merk14, KurtEtal17} provide a similar
``isolated'' experience. Although containerized environments do not
specify hardware-level operations, they are typically packaged with a
complete operating system, in addition to a complete copy of Python
and any relevant package dependencies. Virtual
environments~\cite[e.g.,][]{Anac12} also provide a computing
environment that is largely separated from the user's main
environment. They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproduce an operating system for the runtime environment. Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees---at the hardware level, operating
system level, and Python environment level, respectively) that the
relevant code will run similarly for different users. However, each
of these systems also relies on additional software that can be
resource intensive or burdensome to install or configure.

We designed \texttt{davos} to occupy a ``sweet spot'' between these
extremes.  \texttt{davos} is a notebook-installable package that adds
functionality to the default notebook experience. Like standard
Jupyter notebooks, \texttt{davos}-enhanced notebooks allows
researchers to include text, executable code, and media within a
single file. No further setup or installation is required, beyond
what is needed to run standard Jupyter notebooks. And like virtual
environments, \texttt{davos} provides a convenient mechanism for fully
specifying (and installing, as needed) a complete set of Python
dependencies, including package versions.

>>>>

  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments, to containers, to virtual
  machines.  Each of these existing solutions requires installing or
  setting up a system for running the desired code, increasing the 
  complexity and time cost of sharing or engaging with reproducible 
  science. Here, we propose a lighter-weight solution:  the
  \texttt{davos} library.  When used in combination with a
  notebook-based Python project, \texttt{davos} provides a
  mechanism for specifying (and automatically installing) the correct
  versions of the project's dependencies.  The \texttt{davos}
  library further ensures that those packages and specific versions are used every time the
  notebook's code is executed.  This enables researchers to share a complete
  reproducible copy of their code within a single Jupyter notebook file.

@@

Within the Python~\cite{vanR95} community, external
packages that are published in the most popular
repositories~\cite{Pyth03, cond15} are associated with version numbers
and tags that allow users to guarantee they are installing
exactly the same code across different computing environments \cite{CoghStuf13}.
While it is \textit{possible} to manually install the intended
version of every dependency of a Python script or package, manually
tracking down those dependencies can impose a substantial burden on 
the user and create room for mistakes and inconsistencies. Further, when dependency versions 
are left unspecified, replicating the original computing environment 
becomes difficult or impossible.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code.  Scripts may reference external libraries, but
  those libraries must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those libraries were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.).  When the \textbf{\texttt{davos} library} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install any required external libraries (at their
  correct versions, when specified).  \textbf{Virtual environments}
  allow users to install an isolated copy of Python and all required 
  dependencies. This typically entails distributing a configuration file
  (e.g., a \texttt{pyproject.toml}~\cite{CannEtal16} or \texttt{environment.yml})
  that specifies all project dependencies (including version numbers of 
  external libraries) alongside the primary code base. Users can then install
  a third-praty tool~\cite[e.g.,][]{Anac12, Eust19} to read the file and build the environment.
  \textbf{Containers} provide a means of defining an isolated
  environment that includes a complete operating system (independent
  of the user's operating system), in addition to (optionally)
  specifying a virtual environment or other configurations needed to
  provide the necessary computing environment.  Containers are
  typically defined using specification files (e.g., a plain-text
  \texttt{Dockerfile}) that instruct the virtualization engine
  regarding how to build the containerized environment.  \textbf{Virtual
    machines} provide a complete hardware-level simulation of the
  computing environment.  In addition to simulating specific hardware,
  virtual machines (typically specified using binary images files)
  must also define operating system-level properties of the computing
  environment.  Systems to the left of the blue vertical line entail
  sharing individual files, with no additional installation or
  configuration needed to run the target code.  Systems to the right
  of the red vertical line support precise control over dependencies
  and versioning.  Notebooks enhanced using the \texttt{davos} library
  are easily shareable and require minimal setup costs, while also
  facilitating high reproducibility by enabling precise control over
  project dependencies.}
\label{fig:code-sharing}
\end{figure}

Computational researchers and other programmers have developed a broad
set of approaches and tools to facilitate code sharing and
reproducible outcomes (Fig.~\ref{fig:code-sharing}). At one
extreme, simply distributing a set of Python scripts (\texttt{.py} files) may
enable others to use or gain insights into the relevant work. Because
Python is installed by default on most modern operating systems, for
some projects, this may be sufficient. Another popular approach
entails creating Jupyter
notebooks~\cite{KluyEtal16} that comprise a mix of text, executable
code, and embedded media. Notebooks may call or import external
scripts or libraries---even intersperse snippets of other programming 
or markup lang\-uages---in order to provide a more compact and readable
experience for users. Both of these systems (Python scripts and
notebooks) provide a convenient means of sharing code, with the
caveat that they do not specify the computing environment in which the
code is executed. Therefore the functionality of code shared using
these systems cannot be guaranteed across different users or setups.

At another extreme, virtual machines~\cite{Gold74, AltiEtal05,
 Rose99} provide a hardware-level simulation of the desired system.
Virtual machines are typically isolated such
that installing or running software on a virtual machine does not
impact the user's primary operating system or computing environment.
Containers~\cite[e.g.,][]{Merk14, KurtEtal17} provide a similar
``isolated'' experience. Although containerized environments do not
specify hardware-level operations, they are typically packaged with a
complete operating system, in addition to a complete copy of Python
and any relevant package dependencies. Virtual
environments~\cite[e.g.,][]{Anac12, Eust19} also provide a computing
environment that is largely separated from the user's main
environment. They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproduce an operating system for the runtime environment. Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees---at the hardware level, operating
system level, and Python environment level, respectively) that the
relevant code will run similarly for different users. However, each
of these systems also relies on additional software that can be
complex or resource-intensive to install and use, creating potential 
barriers to both contributing to and taking advantage of open science 
resources.

We designed \texttt{davos} to occupy a ``sweet spot'' between these
extremes.  \texttt{davos} is a notebook-installable package that adds
functionality to the default notebook experience. Like standard
Jupyter notebooks, \texttt{davos}-enhanced notebooks allow
researchers to include text, executable code, and media within a
single file. No further setup or installation is required, beyond
what is needed to run standard Jupyter notebooks. And like virtual
environments, \texttt{davos} provides a convenient mechanism for fully
specifying (and installing, as needed) a complete set of Python
dependencies, including package versions.
